---
title: "Illustration: The Amsterdam Chess Test"
author: "Ruwan Karunanayaka"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustration: The Amsterdam Chess Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Overview

This vignette reproduces the Amsterdam Chess Test illustration from
Section 8.2 of Karunanayaka (2026). The analysis demonstrates three
features of the DGT framework on a single psychometric instrument:

1. Response-scale attenuation for lognormal response times
   (Theorem 1 of the paper).
2. Information loss under Bernoulli sampling for item accuracy
   (Theorem 5).
3. The three-ICC ordering applied to both objects of measurement
   (persons and items) on the same test (Theorem 6).

The dataset is `AmsterdamChess` from the **LNIRT** package: 259 chess
players responding to 40 chess-tactics items, with both response times
(seconds) and correctness scores (0/1). It was collected by van der
Maas and Wagenmakers (2005) and is distributed under an open license.

## Preparation

```{r}
library(LNIRT)
library(brms)
library(dgt)
library(dplyr)
library(tidyr)

data(AmsterdamChess)
act <- AmsterdamChess
```

Three participants (rows 147, 201, 209) have no recorded responses and
are dropped. Missing response times are coded 10000 and missing
accuracy scores are coded 9; both are recoded to `NA`.

```{r}
rt_cols <- grep("^RT([0-9]+)$", names(act), value = TRUE)
y_cols  <- grep("^Y([0-9]+)$",  names(act), value = TRUE)

rt_wide <- act[, rt_cols]; rt_wide[rt_wide == 10000] <- NA
y_wide  <- act[, y_cols];  y_wide[y_wide  == 9]     <- NA

keep <- !apply(rt_wide, 1, function(r) all(is.na(r)))
rt_wide <- rt_wide[keep, ]
y_wide  <- y_wide[keep, ]
```

Reshape to long format with explicit `person_id` and `item_id`
grouping factors:

```{r}
rt_long <- rt_wide |>
  as.data.frame() |>
  mutate(person_id = factor(seq_len(nrow(rt_wide)))) |>
  pivot_longer(all_of(rt_cols), names_to = "item_id",
               values_to = "rt_sec") |>
  filter(!is.na(rt_sec), rt_sec > 0) |>
  mutate(item_id = factor(item_id, levels = rt_cols),
         log_rt  = log(rt_sec))

y_long <- y_wide |>
  as.data.frame() |>
  mutate(person_id = factor(seq_len(nrow(y_wide)))) |>
  pivot_longer(all_of(y_cols), names_to = "item_id",
               values_to = "correct") |>
  filter(!is.na(correct)) |>
  mutate(item_id = factor(item_id, levels = y_cols),
         correct = as.integer(correct))
```

## Response-time model

Fit a lognormal crossed random-effects model:

```{r}
rt_fit <- brm(
  log_rt ~ 1 + (1 | person_id) + (1 | item_id),
  data   = rt_long,
  family = gaussian(),
  prior  = c(prior(normal(4, 2),       class = "Intercept"),
             prior(student_t(3, 0, 1), class = "sd"),
             prior(student_t(3, 0, 1), class = "sigma")),
  chains = 4, cores = 4, iter = 4000, warmup = 1000,
  seed   = 20260418, backend = "cmdstanr"
)
```

### DGT quantities with persons as object

```{r}
icc_rt_person <- dgt_icc(rt_fit, person_group = "person_id")
print(icc_rt_person)
```

Expected output (medians, 95% CIs):

```
ICC_eta (link-scale)        0.173  [0.133, 0.216]
ICC_Y   (response-scale)    0.136  [0.097, 0.176]
ICC_I   (information)       0.173  [0.133, 0.216]
Overestimation (O)          1.28   [1.22, 1.39]
```

### DGT quantities with items as object

Pass `person_group = "item_id"` to treat items as the object of
measurement. The helper is generic in the grouping factor: whatever
random effect is named there becomes the signal, and all other
random effects plus the residual become error.

```{r}
icc_rt_item <- dgt_icc(rt_fit, person_group = "item_id")
print(icc_rt_item)
```

Expected:

```
ICC_eta                     0.426  [0.324, 0.546]
ICC_Y                       0.358  [0.273, 0.458]
ICC_I                       0.426  [0.324, 0.546]
Overestimation (O)          1.19   [1.18, 1.20]
```

Because the lognormal model has an invertible deterministic link
(\eqn{W = \log Y}), Theorem 5 (transformation invariance) gives
\eqn{\icc_I = \icc_\eta} exactly for both objects. The response-scale
ICC is strictly lower than the link-scale ICC, and in both rows the
95% credible interval for the overestimation ratio excludes 1.

### D-study

```{r}
dgt_required_n(rt_fit, target = 0.80, person_group = "person_id")
#> n* (link-scale)   21
#> n* (response-scale)  26

dgt_required_n(rt_fit, target = 0.80, person_group = "item_id")
#> n* (link-scale)   6
#> n* (response-scale)  8
```

## Accuracy model

Fit a Bernoulli crossed random-effects model to the 0/1 correctness
data:

```{r}
acc_fit <- brm(
  correct ~ 1 + (1 | person_id) + (1 | item_id),
  data   = y_long,
  family = bernoulli(link = "logit"),
  prior  = c(prior(normal(0, 2),       class = "Intercept"),
             prior(student_t(3, 0, 1), class = "sd")),
  chains = 4, cores = 4, iter = 4000, warmup = 1000,
  seed   = 20260418, backend = "cmdstanr"
)
```

### Information loss under Bernoulli sampling

New in `dgt` 0.2.0: `dgt_info_icc()` now handles Bernoulli and
binomial families. The function estimates
\eqn{I(\nu_{\mathrm{obj}}; Y)} by nested Monte Carlo and returns the
information ICC, the logit-scale ICC, and their gap. By Theorem 5 the
gap is strictly positive.

```{r}
info_acc_person <- dgt_info_icc(acc_fit, person_group = "person_id")
print(info_acc_person)
```

Expected:

```
I(nu; Y) (mutual information, nats)     0.056  [0.039, 0.074]
ICC_I   (information)                    0.105  [0.076, 0.138]
ICC_eta (logit-scale)                    0.160  [0.118, 0.205]
Gap (ICC_eta - ICC_I)                    0.055  [0.038, 0.074]
```

```{r}
info_acc_item <- dgt_info_icc(acc_fit, person_group = "item_id")
print(info_acc_item)
```

Expected:

```
I(nu; Y) (mutual information, nats)     0.189  [0.143, 0.249]
ICC_I                                    0.315  [0.248, 0.392]
ICC_eta                                  0.480  [0.381, 0.596]
Gap                                      0.165  [0.122, 0.212]
```

The information loss is substantial in both cases: about a third of
the logit-scale between-object variance is structurally unrecoverable
from the discrete 0/1 observations, consistent with the data
processing inequality applied to the Bernoulli sampling step.

## Interpretation

The ACT illustrates all parts of Theorem 6 on a single instrument:

- Response times: `ICC_Y < ICC_eta = ICC_I` for both persons and
  items (Theorem 6(a), continuous invertible link).
- Accuracy: `ICC_I < ICC_eta` for both persons and items (Theorem
  6(b), discrete sampling).

The contrast with the lexical decision illustration (Section 8.1 of
the paper) is instructive. Lexical decision RTs have
\eqn{\sigma^2_\eta \approx 0.12} and \eqn{\mathcal{O} \approx 1.02}:
the response-scale correction is negligible because the total
log-scale variance is small. Chess RTs have
\eqn{\sigma^2_\eta = 0.56} and \eqn{\mathcal{O} = 1.28}: the
correction is meaningful. The correction magnitude depends on total
log-scale variance, not on the ICC itself.

## References

Fox, J.-P., Klotzke, K., & Simsek, A. S. (2023). R-package LNIRT for
joint modeling of response accuracy and times. *PeerJ Computer
Science*, 9, e1232.

Karunanayaka, R. (2026). Distributional Generalizability Theory:
Reliability for Non-Gaussian Measurements. *Psychometrika*,
forthcoming.

van der Maas, H. L. J., & Wagenmakers, E.-J. (2005). A psychometric
analysis of chess expertise. *American Journal of Psychology*,
118(1), 29–60.
