---
title: "Introduction to Distributional Generalizability Theory"
author: "[Authors]"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Distributional Generalizability Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## The Problem

Classical Generalizability Theory defines reliability through the ICC:

$$\text{ICC} = \frac{\sigma^2_p}{\sigma^2_p + \sigma^2_{\text{error}}}$$

This assumes Gaussian measurements. When your data are lognormal (e.g., 
expected goals, reaction times), Poisson (shot counts), or zero-inflated 
(hurdle models), this ICC **overestimates reliability** on the scale you 
actually interpret.

The `dgt` package computes the correct response-scale reliability from any
`brms` model posterior.

## Installation

```{r}
remotes::install_github("rckaruna/dgt")
```

## Example 1: Lognormal Measurements

### Simulate data

```{r}
library(dgt)
library(brms)

set.seed(42)
n_players <- 100
n_matches <- 20

# True variance components (log scale)
sigma_p <- 0.55   # Person SD
sigma_e <- 0.75   # Error SD (occasions + residual)

# Simulate
player_effects <- rnorm(n_players, 0, sigma_p)
dat <- expand.grid(player = 1:n_players, match = 1:n_matches)
dat$player_effect <- player_effects[dat$player]
dat$log_xG <- 0.5 + dat$player_effect + rnorm(nrow(dat), 0, sigma_e)
dat$xG <- exp(dat$log_xG)

# True values:
# ICC_eta = 0.55^2 / (0.55^2 + 0.75^2) = 0.349
# ICC_Y   = (exp(0.3025)-1) / (exp(0.865)-1) = 0.272
```

### Fit model

```{r}
fit <- brm(
  log(xG) ~ 1 + (1 | player),
  data = dat,
  family = gaussian(),
  chains = 4, iter = 2000,
  cores = 4
)
```

### Compute DGT reliability

```{r}
# All three ICCs
result <- dgt_icc(fit, person_group = "player")
print(result)

# --- Distributional Generalizability Theory ---
# Family: gaussian
# 
# ICC Estimates:
#   ICC_eta (link-scale)             0.349  [0.280, 0.421]
#   ICC_Y (response-scale)           0.272  [0.209, 0.340]
#   ICC_I (information)              0.349  [0.280, 0.421]
#   Overestimation (O)               1.28   [1.22, 1.35]
```

**Key result**: The link-scale ICC (0.349) overestimates the response-scale 
ICC (0.272) by a factor of 1.28×. Classical G-theory says this measurement 
is more reliable than it actually is.

### D-study

```{r}
ds <- dgt_dstudy(fit, n_grid = 1:40, person_group = "player")
plot(ds, target = 0.80)
```

This shows three curves:

- **Link-scale** (what classical G-theory reports): reaches 0.80 faster
- **Response-scale (arithmetic mean)**: the correct curve for practitioners
- **Response-scale (geometric mean)**: higher, but targets the median

```{r}
# Required occasions
dgt_required_n(fit, target = 0.80, person_group = "player")
```

### Overestimation analysis

```{r}
oe <- dgt_overestimation(fit, person_group = "player")
print(oe)

# --- DGT Overestimation Analysis (Theorem 6) ---
# 
#   Overestimation ratio (O)         1.28x  [ 1.22,  1.35]
#   D-study ratio (D)                1.38x  [ 1.29,  1.49]
```

The D-study ratio (1.38×) is *worse* than the ICC overestimation (1.28×).
This always holds (Proposition 15).

### ICC posterior densities

```{r}
plot(result)
```


## Example 2: Hurdle-Lognormal Measurements

For measurements with structural zeros (e.g., xG where some players don't
shoot in a match):

```{r}
# Fit hurdle model
fit_hurdle <- brm(
  bf(xG ~ 1 + (1 | player),
     hu ~ 1 + (1 | player)),
  data = dat,
  family = hurdle_lognormal(),
  chains = 4, iter = 2000
)

# Three-part reliability decomposition
result_h <- dgt_icc(fit_hurdle, person_group = "player")
print(result_h)

# --- Distributional Generalizability Theory ---
# Family: hurdle_lognormal
# 
# ICC Estimates:
#   ICC_Z (engagement)               0.146  [0.082, 0.224]
#   ICC_Y* (intensity)               0.272  [0.205, 0.345]
#   ICC_comp (composite)             0.199  [0.140, 0.268]
# 
# Variance Decomposition:
#   V1 (binary noise)                0.4521  (35.2%)
#   V2 (continuous noise)            0.3812  (29.7%)
#   V3 (engagement signal)           0.0891  ( 6.9%)
#   V4 (intensity signal)            0.2104  (16.4%)
#   V5 (interaction signal)          0.1512  (11.8%)
```

### Variance decomposition diagnostic

```{r}
vd <- dgt_variance(fit_hurdle, person_group = "player")
print(vd)

# Bottleneck: Binary engagement process 
#   (increase n_m for more Bernoulli trials)
```

### Information decomposition

```{r}
info <- dgt_info_icc(fit_hurdle, person_group = "player")
print(info)

# --- DGT Information ICC (Theorems 4-5) ---
# Method: monte_carlo
# 
#   I_engage                                    0.078  [0.040, 0.128]
#   I_intensity (per occasion, given Z=1)       0.347  [0.278, 0.421]
#   I_total                                     0.250  [0.188, 0.318]
#   ICC_I                                       0.394  [0.313, 0.471]
```


## Example 3: Poisson Counts

For count data (e.g., number of shots per match):

```{r}
# Simulate Poisson counts
dat$shots <- rpois(nrow(dat), exp(1.5 + player_effects[dat$player] * 0.3))

fit_pois <- brm(
  shots ~ 1 + (1 | player),
  data = dat,
  family = poisson(),
  chains = 4, iter = 2000
)

result_p <- dgt_icc(fit_pois, person_group = "player")
print(result_p)
```


## Example 4: Binary Choices and a Decision Study

A paired-choice discrete choice experiment (DCE) records, for each
respondent and each choice task, which of two profiles was chosen. The
response is Bernoulli. A decision study asks how many tasks per
respondent are needed before a respondent-level quantity (the decrement
a respondent attaches to an attribute level, say) is estimated to a
stated reliability. On the logit scale this is the classical
Spearman-Brown calculation with the $\pi^2/3$ residual convention. On
the probability scale, which is where choice probabilities and value
sets live, reliability is lower and the required number of tasks is
larger. Both curves come from one fit.

### Simulate a paired-choice design

300 respondents and 21 tasks each. The respondent SD (1.0) and the task
SD (0.6) on the logit scale are modelling assumptions, not estimates
from any published study.

```{r}
set.seed(1)
n_resp  <- 300
n_task  <- 21
sigma_r <- 1.0
sigma_t <- 0.6

u <- rnorm(n_resp, 0, sigma_r)
v <- rnorm(n_task, 0, sigma_t)

dce <- expand.grid(resp = 1:n_resp, task = 1:n_task)
dce$eta <- u[dce$resp] + v[dce$task]
dce$y   <- rbinom(nrow(dce), 1, plogis(dce$eta))
```

### Fit

```{r}
fit_dce <- brm(
  y ~ 1 + (1 | resp) + (1 | task),
  data = dce,
  family = bernoulli(),
  chains = 4, iter = 2000,
  cores = 4
)
```

### Decision study

`person_group = "resp"` makes the respondent the object of measurement;
the task intercept is treated as a facet and integrated out. `K` and
`K_facet` set the outer and inner Monte Carlo sizes for the
response-scale curve.

```{r}
ds_dce <- dgt_dstudy(fit_dce, n_grid = 1:60, person_group = "resp",
                     K = 2000, K_facet = 300, seed = 1)
print(ds_dce)

# --- DGT D-Study ---
# Family: bernoulli
#
# Required occasions for target reliability:
#  threshold link_scale response_scale
#        0.7          8             12
#        0.8         13             20
#        0.9         29             44

plot(ds_dce, target = 0.80)
```

The response-scale curve reaches 0.70 at 12 tasks and 0.80 at 20; the
link-scale curve reaches the same targets at 8 and 13. Reading off the
design's own 21 tasks:

```{r}
subset(ds_dce$curves, n == 21, select = c(type, median, lower, upper))

#             type median lower upper
#       link-scale  0.869  ...   ...
#   response-scale  0.814  ...   ...
```

Two things to note. The gap between the curves is not sampling error;
response-scale reliability is below link-scale reliability for every
`n` (the ordering `response <= link` is checked in the package tests).
And the answer depends on the respondent SD: halving it to 0.5 pushes
the response-scale 0.80 target beyond 60 tasks, so a decision study
also states what respondent heterogeneity a recommended task count
presupposes.

### Heterogeneous attendance

If a fraction of respondents attend to the attribute with reduced
signal (attribute non-attendance in the DCE literature), the population
is a mixture over respondents and the variance components of the
estimated decrement change. Simulating 40% of respondents at half
signal and refitting moves the response-scale requirements to 18, 30,
and beyond 60 tasks for 0.70, 0.80, and 0.90. The direction is general;
the size depends on the mixture. Non-attendance therefore enters the
decision study as part of the population being generalized over rather
than as a data-cleaning step applied before it.

### Information curve

`info = TRUE` adds the information-theoretic curve of Theorems 4-5. It
is the expensive curve and sits at or below the link-scale curve.

```{r}
ds_dce_info <- dgt_dstudy(fit_dce, n_grid = 1:60, person_group = "resp",
                          K = 2000, K_facet = 300, info = TRUE, seed = 1)
```

The facet is pooled in this release: respondents and tasks are not
allocated separately. That extension is planned.


## The Three-ICC Framework

DGT provides three complementary reliability measures:

| Measure | Question answered | When to use |
|---|---|---|
| ICC_Y | How correlated are my observations? | Reporting reliability |
| ICC_eta | What does classical G-theory say? | Comparison with literature |
| ICC_I | How much information per observation? | Design decisions |

For Gaussian data, all three are equal. For non-Gaussian data:

- **Lognormal**: ICC_Y < ICC_eta = ICC_I (correlation distortion only)
- **Poisson**: ICC_I < ICC_eta (genuine information loss)
- **Hurdle**: ICC_Y < ICC_I < ICC_eta (both problems)


## Practical Workflow

```{r}
# 1. Fit your GLMM in brms
fit <- brm(...)

# 2. Compute all ICCs
icc_result <- dgt_icc(fit, person_group = "your_person_var")

# 3. Check overestimation (lognormal only)
oe <- dgt_overestimation(fit, person_group = "your_person_var")

# 4. Run D-study
ds <- dgt_dstudy(fit, person_group = "your_person_var")
plot(ds, target = 0.80)

# 5. Determine required observations
dgt_required_n(fit, target = 0.80, person_group = "your_person_var")

# 6. For hurdle models: identify the bottleneck
vd <- dgt_variance(fit, person_group = "your_person_var")
```


## Mathematical Foundation

The package implements six theorems from Distributional Generalizability Theory:

- **Theorem 1**: $\text{ICC}_Y^{\text{LN}} = \frac{\exp(\sigma^2_p)-1}{\exp(\sigma^2_\eta)-1}$
- **Theorem 1a**: $\text{ICC}_Y < \text{ICC}_\eta$ always (classical G-theory overestimates)
- **Theorem 2**: D-study = Spearman-Brown with ICC_Y (same formula, different input)
- **Theorem 3**: Hurdle ICC with five-component variance decomposition
- **Theorem 5**: $I(\nu; Y) = I(\nu; g(Y))$ — information is invariant under invertible links
- **Theorem 6**: ICC_Y < ICC_eta = ICC_I, with overestimation unbounded as skewness grows

See the paper: "Distributional Generalizability Theory: Reliability for Non-Gaussian Measurements."
