---
title: "Introduction to Distributional Generalizability Theory"
author: "[Authors]"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Distributional Generalizability Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## The Problem

Classical Generalizability Theory defines reliability through the ICC:

$$\text{ICC} = \frac{\sigma^2_p}{\sigma^2_p + \sigma^2_{\text{error}}}$$

This assumes Gaussian measurements. When your data are lognormal (e.g., 
expected goals, reaction times), Poisson (shot counts), or zero-inflated 
(hurdle models), this ICC **overestimates reliability** on the scale you 
actually interpret.

The `dgt` package computes the correct response-scale reliability from any
`brms` model posterior.

## Installation

```{r}
remotes::install_github("username/dgt")
```

## Example 1: Lognormal Measurements

### Simulate data

```{r}
library(dgt)
library(brms)

set.seed(42)
n_players <- 100
n_matches <- 20

# True variance components (log scale)
sigma_p <- 0.55   # Person SD
sigma_e <- 0.75   # Error SD (occasions + residual)

# Simulate
player_effects <- rnorm(n_players, 0, sigma_p)
dat <- expand.grid(player = 1:n_players, match = 1:n_matches)
dat$player_effect <- player_effects[dat$player]
dat$log_xG <- 0.5 + dat$player_effect + rnorm(nrow(dat), 0, sigma_e)
dat$xG <- exp(dat$log_xG)

# True values:
# ICC_eta = 0.55^2 / (0.55^2 + 0.75^2) = 0.349
# ICC_Y   = (exp(0.3025)-1) / (exp(0.865)-1) = 0.272
```

### Fit model

```{r}
fit <- brm(
  log(xG) ~ 1 + (1 | player),
  data = dat,
  family = gaussian(),
  chains = 4, iter = 2000,
  cores = 4
)
```

### Compute DGT reliability

```{r}
# All three ICCs
result <- dgt_icc(fit, person_group = "player")
print(result)

# --- Distributional Generalizability Theory ---
# Family: gaussian
# 
# ICC Estimates:
#   ICC_eta (link-scale)             0.349  [0.280, 0.421]
#   ICC_Y (response-scale)           0.272  [0.209, 0.340]
#   ICC_I (information)              0.349  [0.280, 0.421]
#   Overestimation (O)               1.28   [1.22, 1.35]
```

**Key result**: The link-scale ICC (0.349) overestimates the response-scale 
ICC (0.272) by a factor of 1.28×. Classical G-theory says this measurement 
is more reliable than it actually is.

### D-study

```{r}
ds <- dgt_dstudy(fit, n_grid = 1:40, person_group = "player")
plot(ds, target = 0.80)
```

This shows three curves:

- **Link-scale** (what classical G-theory reports): reaches 0.80 faster
- **Response-scale (arithmetic mean)**: the correct curve for practitioners
- **Response-scale (geometric mean)**: higher, but targets the median

```{r}
# Required occasions
dgt_required_n(fit, target = 0.80, person_group = "player")
```

### Overestimation analysis

```{r}
oe <- dgt_overestimation(fit, person_group = "player")
print(oe)

# --- DGT Overestimation Analysis (Theorem 6) ---
# 
#   Overestimation ratio (O)         1.28x  [ 1.22,  1.35]
#   D-study ratio (D)                1.38x  [ 1.29,  1.49]
```

The D-study ratio (1.38×) is *worse* than the ICC overestimation (1.28×).
This always holds (Proposition 15).

### ICC posterior densities

```{r}
plot(result)
```


## Example 2: Hurdle-Lognormal Measurements

For measurements with structural zeros (e.g., xG where some players don't
shoot in a match):

```{r}
# Fit hurdle model
fit_hurdle <- brm(
  bf(xG ~ 1 + (1 | player),
     hu ~ 1 + (1 | player)),
  data = dat,
  family = hurdle_lognormal(),
  chains = 4, iter = 2000
)

# Three-part reliability decomposition
result_h <- dgt_icc(fit_hurdle, person_group = "player")
print(result_h)

# --- Distributional Generalizability Theory ---
# Family: hurdle_lognormal
# 
# ICC Estimates:
#   ICC_Z (engagement)               0.146  [0.082, 0.224]
#   ICC_Y* (intensity)               0.272  [0.205, 0.345]
#   ICC_comp (composite)             0.199  [0.140, 0.268]
# 
# Variance Decomposition:
#   V1 (binary noise)                0.4521  (35.2%)
#   V2 (continuous noise)            0.3812  (29.7%)
#   V3 (engagement signal)           0.0891  ( 6.9%)
#   V4 (intensity signal)            0.2104  (16.4%)
#   V5 (interaction signal)          0.1512  (11.8%)
```

### Variance decomposition diagnostic

```{r}
vd <- dgt_variance(fit_hurdle, person_group = "player")
print(vd)

# Bottleneck: Binary engagement process 
#   (increase n_m for more Bernoulli trials)
```

### Information decomposition

```{r}
info <- dgt_info_icc(fit_hurdle, person_group = "player")
print(info)

# --- DGT Information ICC (Theorems 4-5) ---
# Method: monte_carlo
# 
#   I_engage                                    0.078  [0.040, 0.128]
#   I_intensity (per occasion, given Z=1)       0.347  [0.278, 0.421]
#   I_total                                     0.250  [0.188, 0.318]
#   ICC_I                                       0.394  [0.313, 0.471]
```


## Example 3: Poisson Counts

For count data (e.g., number of shots per match):

```{r}
# Simulate Poisson counts
dat$shots <- rpois(nrow(dat), exp(1.5 + player_effects[dat$player] * 0.3))

fit_pois <- brm(
  shots ~ 1 + (1 | player),
  data = dat,
  family = poisson(),
  chains = 4, iter = 2000
)

result_p <- dgt_icc(fit_pois, person_group = "player")
print(result_p)
```


## The Three-ICC Framework

DGT provides three complementary reliability measures:

| Measure | Question answered | When to use |
|---|---|---|
| ICC_Y | How correlated are my observations? | Reporting reliability |
| ICC_eta | What does classical G-theory say? | Comparison with literature |
| ICC_I | How much information per observation? | Design decisions |

For Gaussian data, all three are equal. For non-Gaussian data:

- **Lognormal**: ICC_Y < ICC_eta = ICC_I (correlation distortion only)
- **Poisson**: ICC_I < ICC_eta (genuine information loss)
- **Hurdle**: ICC_Y < ICC_I < ICC_eta (both problems)


## Practical Workflow

```{r}
# 1. Fit your GLMM in brms
fit <- brm(...)

# 2. Compute all ICCs
icc_result <- dgt_icc(fit, person_group = "your_person_var")

# 3. Check overestimation (lognormal only)
oe <- dgt_overestimation(fit, person_group = "your_person_var")

# 4. Run D-study
ds <- dgt_dstudy(fit, person_group = "your_person_var")
plot(ds, target = 0.80)

# 5. Determine required observations
dgt_required_n(fit, target = 0.80, person_group = "your_person_var")

# 6. For hurdle models: identify the bottleneck
vd <- dgt_variance(fit, person_group = "your_person_var")
```


## Mathematical Foundation

The package implements six theorems from Distributional Generalizability Theory:

- **Theorem 1**: $\text{ICC}_Y^{\text{LN}} = \frac{\exp(\sigma^2_p)-1}{\exp(\sigma^2_\eta)-1}$
- **Theorem 1a**: $\text{ICC}_Y < \text{ICC}_\eta$ always (classical G-theory overestimates)
- **Theorem 2**: D-study = Spearman-Brown with ICC_Y (same formula, different input)
- **Theorem 3**: Hurdle ICC with five-component variance decomposition
- **Theorem 5**: $I(\nu; Y) = I(\nu; g(Y))$ — information is invariant under invertible links
- **Theorem 6**: ICC_Y < ICC_eta = ICC_I, with overestimation unbounded as skewness grows

See the paper: "Distributional Generalizability Theory: Reliability for Non-Gaussian Measurements."
